---
title: "hist"
---

```{r}
library(tidyverse)
```


# Histograms

## Purpose of Histograms

- Histograms are used to visualize the distribution of a single variable
- They are used for continuous variables (e.g., income, age, etc.)
- A *continuous* variable is one that can take on any value within a range (e.g., 0.5, 1.2, 3.7, etc.)
- A *discrete* variable is one that can only take on certain values (e.g., 1, 2, 3, etc.)
- x-axis represents value of variable of interest
- y-axis represents the frequency of that value

## Example

```{r}
#| label: histogram1

# load data
dem_women <- read_csv("data/dem_women.csv")

# filter to 2022
dem_women_2022 <- dem_women |>
  filter(year == 2022) 

# create histogram
ggplot(dem_women_2022, aes(x = flfp)) +
  geom_histogram(fill = "steelblue") + 
  labs(
    x = "Percentage of Working Aged Women in Labor Force",
    y = "Number of Countries",
    title = "Female labor force participation rates, 2022",
    caption = "Source: World Bank"
    ) + theme_minimal()
```

## Histogram Code

<br>

```{r}
#| label: histogram2
#| echo: true
#| eval: false
#| code-line-numbers: "|1-2|4-6|8-10"

# load data
dem_women <- read_csv("data/dem_women.csv")

# filter to 2022
dem_women_2022 <- dem_women |>
  filter(year == 2022) 

# create histogram
ggplot(dem_women_2022, aes(x = flfp)) +
  geom_histogram(fill = "steelblue") + 
  labs(
    x = "Percentage of Working Aged Women in Labor Force",
    y = "Number of Countries",
    title = "Female labor force participation rates, 2022",
    caption = "Source: World Bank"
    ) + theme_minimal()
```

## Change Number of Bins

<br>

Change number of bins (bars) using `bins` or `binwidth` arguments (default number of bins = 30):

<br>

```{r}
#| label: histogram3
#| echo: true
#| eval: false
#| code-line-numbers: "2"

ggplot(dem_women_2022, aes(x = flfp)) +
  geom_histogram(bins = 50, fill = "steelblue") + 
  labs(
    x = "Percentage of Working Aged Women in Labor Force",
    y = "Number of Countries",
    title = "Female labor force participation rates, 2022",
    caption = "Source: World Bank"
    ) + theme_minimal()
```

##

At 50 bins...

```{r}
#| label: histogram4

ggplot(dem_women_2022, aes(x = flfp)) +
  geom_histogram(bins = 50, fill = "steelblue") + 
  labs(
    x = "Percentage of Working Aged Women in Labor Force",
    y = "Number of Countries",
    title = "Female labor force participation rates, 2022",
    caption = "Source: World Bank"
    ) + theme_minimal()
```

##

At 100 bins...probably too many!

```{r}
#| label: histogram5

ggplot(dem_women_2022, aes(x = flfp)) +
  geom_histogram(bins = 100, fill = "steelblue") + 
  labs(
    x = "Percentage of Working Aged Women in Labor Force",
    y = "Number of Countries",
    title = "Female labor force participation rates, 2022",
    caption = "Source: World Bank"
    ) + theme_minimal()
```

## 

<br>

Using `binwidth` instead of `bins`... 

<br>


```{r}
#| label: histogram6
#| echo: true
#| eval: false
#| code-line-numbers: "2"

ggplot(dem_women_2022, aes(x = flfp)) +
  geom_histogram(binwidth = 2, fill = "steelblue") + 
  labs(
    x = "Percentage of Working Aged Women in Labor Force",
    y = "Number of Countries",
    title = "Female labor force participation rates, 2022",
    caption = "Source: World Bank"
    ) + theme_minimal()
```

##

Setting `binwidth` to 2...

```{r}
#| label: histogram7

ggplot(dem_women_2022, aes(x = flfp)) +
  geom_histogram(binwidth = 2, fill = "steelblue") + 
  labs(
    x = "Percentage of Working Aged Women in Labor Force",
    y = "Number of Countries",
    title = "Female labor force participation rates, 2022",
    caption = "Source: World Bank"
    ) + theme_minimal()
```

## Change from Count to Density

<br>

```{r}
#| label: histogram8
#| echo: true
#| eval: false
#| code-line-numbers: "1,5"


ggplot(dem_women_2022, aes(after_stat(density), x = flfp)) +
  geom_histogram(fill = "steelblue") + 
  labs(
    x = "Percentage of Working Aged Women in Labor Force",
    y = "Density",
    title = "Female labor force participation rates, 2022",
    caption = "Source: World Bank"
    ) + theme_minimal()
```

## 

Which gives us... 

```{r}
#| label: histogram9

ggplot(dem_women_2022, aes(after_stat(density), x = flfp)) +
  geom_histogram(fill = "steelblue") + 
  labs(
    x = "Percentage of Working Aged Women in Labor Force",
    y = "Density",
    title = "Female labor force participation rates, 2022",
    caption = "Source: World Bank"
    ) + theme_minimal()
```

## Your Turn!

1. Pick a variable that you want to explore the distribution of
2. Make a histogram
    a. Only specify `x = ` in `aes()`
    b. Specify geom as `geom_histogram`
3. Choose color for bars
4. Choose appropriate labels
5. Change number of bins
6. Change from count to density

```{r}
#| label: timer3
countdown::countdown(minutes = 10, id = "timer3")
```

